Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-07 report/emse24.tex
feat(report): emse24 updates

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks. Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility.

% It aligns perfectly with the feedback-driven nature of machine learning development, as it supports rapid prototyping and iterative refinement of models.

% TODO: back the TDD angle not only from scientific literature but also general methodologies used and adopted by the software engineering community (Xtreme programming, Kent Beck)

The predominant form of obtaining feedback in Jupyter notebooks is through the output of code cells. These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations. Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code. This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.

% TODO cite scully and breck and the data validation papers here...
% TODO: why are we doing this? how is the prior work lacking? why is this important?

In Jupyter notebooks, feedback is crucial for making important decisions throughout the machine learning development lifecycle. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. As data, personnel, and system environments inevitably change, undocumented decisions can lead to significant technical debt. This not only prolongs debugging and troubleshooting times but also incurs additional costs and resource expenditures.

In software engineering, tests are employed to document the expected behavior of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. By embedding assertions, developers ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

% TODO introduce our results and key findings...

% and then to tie the two things together: are assertions derived from the output of a cell? we still need to find a way to justify why we only look at assertions derived from visualisations specifically

% By analyzing the outputs and assertions in Jupyter notebooks, we aim to provide insights into the feedback mechanisms used in machine learning development, ultimately proposing strategies to improve the efficiency and robustness of this iterative process.

@@

% building ml systems is highly iterative and experimental. information gained from ml visualisations is used to make design and implementation decisions for the following steps of the ml pipeline. for instance, we may visualise the distribution of our training data and find that it is normally distributed. based on this information, we may opt for a linear regression model which assumes normality in the underlying data. however, such expectations regarding the data may be violated once the ml system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.

% implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ml system are no longer satisfied.

% in contrast to prior work, we approach ml testing from a new perspective. we conduct an empirical analysis of computational notebooks obtained from github, to understand the process of testing ml systems in practice. in particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions). the research questions along with the contributions of this paper are as follows.

@@

\section{Preliminaries}\label{sec:prelim}

Computational notebooks have been ubiquitously adopted by the machine
learning community for developing ML enabled systems. Although
originally intended to promote reproducible software, computational
notebooks are far from being reproducible due to lack of software
engineering best practices such as separation of concern, testing and
versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.

\emph{Psallidas et al} provide an overview of the evolving landscape
of computational notebooks by analysing six million Python notebooks,
two million enterprise Data Science (DS) pipelines, source code and
metadata from over 900 releases of 12 important DS libraries. Their
findings can be used by system builders for improving DS tools and
also by DS practitioners to understand the current trends in
technologies they should focus on. \emph{Pimentel et al} mined 1.4
million notebooks from Github to conduct an empirical study on the
coding practices in computational notebooks. Based on their analysis,
the authors propose guidelines on improving reproducibility of
computational notebooks. To enable future research on computational
notebooks, \emph{Quaranta et al} mine Kaggle and present
\textit{KGTorrent}, a public dataset consisting of approximately 2.5
million Jupyter notebooks written in Python. The dataset also contains
a relational database dump of metadata regarding publicly available
notebooks on Kaggle.

Studies with human subjects have been conducted to gain a deeper
understanding of the challenges faced by ML practitioners when
developing ML models inside notebooks. The results indicate that ML
practitioners work in a highly iterative fashion, often experimenting
with multiple strategies to analyse the data or produce meaningful
visualisations~\cite{kandel2012enterprise, kery2018story,
  liu2019understanding, chattopadhyay2020wrong}. Studies have also
been conducted to understand how practitioners generate, evaluate and
manage alternative hypothesis, visual designs, methods, tools,
algorithms and data sources to arrive at the final
implementation~\cite{liu2019understanding,kandel2012enterprise}. The
findings from these studies can be used as guidelines for improving
existing notebook technologies or designing new ones.

@@

@@

@@

>>>>

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.

Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks. Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

In software engineering, tests are employed to document the expected behavior of code. Failing tests indicate when these expectations are no longer met. Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development. Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks. Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.

% TODO: contributions?
% TODO: key observations?

@@

@@

\section{Background}\label{sec:background}

\subsection{Jupyter Notebooks}

% NOTE use background section from pimentel2019large-scale

% describe the structure of jupyter notebooks (anatomy: cells, outputs); no need to go into too much technical detail regarding the internal JSON structure, just mention that the data is represented in a standardised structure internally

\subsection{Python Abstract Syntax Tree}

% explain the high-level way in which ast module in python works
% explain the anatomy of assert statements and the more complicated way in which asts for print statements are represented in an AST

@@

% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells. These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations. Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code. This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.

@@

\section{Threats to Validity}\label{sec:threats}
TODO

\section{Related Work}\label{sec:related}

\subsection{Feedback in Software Engineering and Machine Learning}

% the role of early adoption of feedback when developing software, agile
% similar concept exists in ML: crisp-dm, mldl, sculley, amershi, sambasivan
% feedback from one part of the pipeline feeds into another,

% start with a very quick review of the workflow: REPL style of development; break code into smaller chunks; faster feedack; iterative style of development
% link to the existing work on notebooks from chi: design, benefits and challenges
% then go into the SE side of software written in notebooks: code quality
% adoption of notebooks by ML community

% start with a quick introduction to the anatomy of a computational notebook: code is broken down into smaller chunks represented as cells, cells can contain code (which are executable and immediately display the result of the execution), or markdown cells used for the purpose of documenting or creating a narrative for collaborative work.

% the move on to the benefits of this style of coding: the ML pipeline contains several components, which can be broken down into smaller, more manageable chunks. These can be individually developed incorporating the rapid feedback provided from the REPL style of coding.

% then summarise the pain-points of working in notebooks: use chattopadhyay2020whats
Computational notebooks have been ubiquitously adopted by the ML community for developing ML enabled systems. Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing, dependenvcy management and versioning~\cite{pimentel2019large-scale,wang2020better,chattopadhyay2020whats}.

\emph{Psallidas et al}~\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries. Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on. \emph{Pimentel et al}~\cite{pimentel2019large-scale} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks. Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks. To enable future research on computational notebooks, \emph{Quaranta et al}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python. The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.

Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks. The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020whats}. Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}. The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.

@@

% NOTE: taxonomy of assertions mapping to V&V tasks in ML
% NOTE: automated tool that adds assertions once the contents of the NB is finalised?
